---
title: "3: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Call packages in R

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # this causes the statement of numeric vector in vector to call up a specific item in square brackets... specifically the third item, and it outputs the number

# Matrices ---- creating a matrix that has two dimensions
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # matrix is a function in R with numbers from 1-20 in order in each colum
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) # this places the numbers in order numerically across a row and not column, and "Byrow = true" shows this population by row
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #
matrix3 # labled the columns and rows with names using the "dimnames" code
matrix1[4, ] # in Matrix one it is calling up row 4
matrix1[ , 3] # in matrix 1 it is calling up column 3 numbers
matrix1[c(12, 14)] # just grabbed where the numbers within the matrix - specificing columns 
matrix1[c(12:14)] # brought up numbers 12-14
matrix1[2:4, 1:3] # called up rows 2-4 and columns 1-3 and their respective numbers

cells <- c(1, 26, 24, 68) #creating a vector of cells
rnames <- c("R1", "R2") #create another character vector for rows
cnames <- c("C1", "C2") # create another character vector for columns
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # this tells R how to create this matrix 2x2 byrows left to right and specificy the "dimnames
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1 #creating a list with a list fuction with a name called Maria, it has a character vector called my numbers, and show the name, : this allows you to do two tasks and only show one answer. this can be useful for multiple different types and sizes of information like a medical record
list1[[2]] #

# Data Frames ---- #each column has a different character and building a datafram from start to finish
d <- c(1, 2, 3, 4) # What type of vector? numeric
e <- c("red", "white", "red", NA) # What type of vector? character
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logical
dataframe1 <- data.frame(d,e,f) # want data included -creates a frame
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # giving rows and columns names

dataframe1[1:2,] # subsetting within dataframe shows rows and only rows you called[rows, column]
dataframe1[c("ID","Passed")] # calls up just the columns within the ID and Passed
dataframe1$ID #datafram calls up IDs column, and $ indicates calling up a specific portion

```
Question: How do the different types of data appear in the Environment tab?

> Answer: Within the environment tab creating matrixies and dataframs show our data and what we are asking it to do, and well as what we assigned it to do within the matrix or dataset. If you clicked the pull down button or on the function it will create a new tab and show an actual chart of your data.

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: Hitting tab picks the first item in your columns list and then calls up all of the data into the column.

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
vector4 <- c(62, 45, 50, 50, 40, 40, 45, 52, 54, 58) 
vector4
vector5 <- c(31, 29, 43, 24, 23, 23, 26, 30, 42, 41)
vector5
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius.

```{r}
vector4 <- c(62, 45, 50, 50, 40, 40, 45, 52, 54, 58) 
vector4-
vector5 <- c(31, 29, 43, 24, 23, 23, 26, 30, 42, 41)
vector5/32
```

Combine your four vectors into a data frame and add informative column names.

```{r}

```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}

```


## Packages

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your libaray already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a package is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.


```{r}
# We will use the packages dplyr and ggplot2 regularly. 
install.packages("dplyr") 
install.packages("ggplot2")
# comment out install commands, use only when needed and re-comment

library(dplyr)
library(ggplot2)

# Some packages are umbrellas under which other packages are loaded
install.packages("tidyverse")
library(tidyverse)
```

Question: What happens in the console when you load a package?

> Answer: It shows what packages you have, and library helps load in packages and keep them updated, and in conflicts tells you that these functions are doing something else


## Tips and Tricks

* Sequential section headers can be created by using at least four -, =, and # characters.

* The command `require(packagename)` will also load a package, but it will not give any error or warning messages if there is an issue.

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 

* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out. 

* Update your packages regularly! 