---
title: "5: Data Exploration"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
fig_width: 5
fig_height: 2.5
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packges
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv")

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") 
class(USGS.flow.data$datetime) #in my USGS folderwe want to assign it as a date and replace the date as year, month, and day

```


## Visualization for Data Exploration

Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar() #aes is aesthetics and we are showing how many counts of the discharge mean approval and create a geom bar graph. We want Approved and Provisional data. It consists of mostly approved fata
```

### Histogram (function: geom_histogram)

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
#gg plot of USGS flow data with aes within the histogram to visualize discharge mean
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean))

#told the width of each bin to be 10 units wide
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

# told to make 20 bins and it decides which works best according to R
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 20)

#sets limitis by starting at 0 and go to 500, + at the end of the line not a , so that r can continue to run more code after
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500))
  
#this shows histogram of gage height
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))
#print(gageplot) to show up in the knitted PDF
```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#make a ggplot and then ask for a geom polygon for gageheight with 50 bins; make another geom polygon to show minimum gage height with 50 bins with a color to decipher it; lty=2 line type 2 creates a dashed line- 3 is a dotted line; created a graph with x axis from 0 to 10
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "pink") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 4) +
  scale_x_continuous(limits = c(0, 10))
#find how many rows or NAs with the summary function
summary(USGS.flow.data$gage.height.max)
summary(USGS.flow.data$gage.height.min)
summary(USGS.flow.data$gage.height.mean)

#frequency polygon to create a geom with x as mean gage height but we want to know whether they are approved or provisional by color & place a legend for the graph at the top. We have few provisional data thats why it is a low line
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin)

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#high outliers for gage height and relative distributions between A and P we can't tell the difference in sample sizes. P has a lot less data than A and less outliers
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#specify x as a continuious varible instead of factor levels. It will give us different distributions across a range of data
ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#violin plot that shows information about sample counts and will show more data with width. You can also place quartiles in there and the ends of the violin are the bottom and top numbers in your dataset
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
              scale = "count")
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data) +
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: To show data in a visual way instead of an excel spreadsheet. It might also help you see correlations in your data that you might not see through excel in a big dataset.

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: That provivisons have a lot less of sampling to look at and less outliers. Approved has a more robust data set with outliers but is heavily weighted to one side.


How can multiple options for data exploration inform our understanding of our data?

> Answer: Some such as the box plot might not always show the differences in datasets where as a violin dataset will show data robustness with width.


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: It is heavily weighted to one side and the approved colum has a lot more data to work with. 


