---
title: "Assignment 5: Data Visualization"
author: "Masha Edmondson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 11 at 1:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse and cowplot packages. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (tidy and gathered) and the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
getwd()
library(tidyverse)
library(ggridges)
library(cowplot)
PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <-
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
NEON_NIWO_Litter_Processed <- read_csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
PeterPaul.nutrients.gathered$sampledate <- as.Date(PeterPaul.nutrients.gathered$sampledate, format = "%Y-%m-%d")

class(PeterPaul.chem.nutrients$sampledate)
class(PeterPaul.chem.nutrients.gathered$sampledate)
class(NEON_NIWO_Litter_Processed$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
mytheme <- theme_classic(base_size = 14) + 
  theme(axis.text = element_text(color = "black"),  
        legend.position = "right") 

theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus by phosphate, with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
TP_PO4_plot <- ggplot(PeterPaul.chem.nutrients, aes(x= tp_ug, y = po4, shape= lakename, color =lakename)) + 
  geom_point(aes(x= tp_ug, y = po4, shape= lakename, color =lakename))+
  geom_smooth(aes(x= tp_ug, y = po4, shape= lakename, color =lakename), method = "lm", se = FALSE, color="black") +
  labs(x = expression(paste("Total Phosphorus (", mu, "g/L)")), y = expression(paste("Phosphate (", mu, "g/L)")), color= "lakename") +
  xlim(0, 150)+
  ylim(0, 50)+
  ggtitle(" Total Phosphorus by Phosphate in Peter and Paul Lake")
print(TP_PO4_plot)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
PeterPaul.temp.plot <-
  ggplot(PeterPaul.chem.nutrients, aes(x= as.factor(month), y= temperature_C, color = lakename))+
  geom_boxplot(varwidth = TRUE)+
  labs(x = "Month", y = "Temperature (C)", color = "Lake Name")+
  theme(legend.position = "top")
print(PeterPaul.temp.plot)

PeterPaul.tp.plot <-
  ggplot(PeterPaul.chem.nutrients, aes(x= as.factor(month), y = tp_ug, color = lakename)) +
  geom_boxplot() +
  theme(legend.position = "none")+
  labs(x = "Month", y = expression(paste("Total Phosphorus (", mu, "g/L)"))) +
  scale_y_continuous(expand = c(0,0))+
  #date_labels = "%B"
print(PeterPaul.tp.plot)

PeterPaul.tn.plot <-
  ggplot(PeterPaul.chem.nutrients, aes(x= as.factor(month), y = tn_ug, color = lakename)) +
  geom_boxplot() +
  theme(legend.position = "none")+
  labs(x = "Month", y = expression(paste("Total Nitrogen (", mu, "g/L)"))) +
  scale_y_continuous(expand = c(0,0))+
  #date_labels = "%B"
print(PeterPaul.tn.plot)

plot_grid(PeterPaul.temp.plot, PeterPaul.tp.plot, PeterPaul.tn.plot, nrows= 3, align = "h", rel_heights = c(1.25, 1)) 

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
Needles.data <- subset(NEON_NIWO_Litter_Processed, functionalGroup == "Needles" )
Needles.plot <- ggplot(subset(NEON_NIWO_Litter_Processed, functionalGroup == "Needles"), aes( x = as.Date(collectDate), y = dryMass, color = nlcdClass))+
  geom_point()+
  labs(x= "", y = expression("Dry Mass (g)"), color= "NLCD CLASS")+
  #scale_color_viridis(option = "magma", direction = -1, end = 0.8)+
  theme(legend.position = "right")
print(Needles.plot)

#7
facets.plot <-
  ggplot(subset(NEON_NIWO_Litter_Processed, functionalGroup == "Needles"), aes(x= collectDate, y= dryMass, color= nlcdClass))+
  labs(x= "", y= expression ("Dry Mass (g)"), color = "NLCD Class")+
  geom_point()+
  theme(legend.position = "right")+
  facet_wrap(vars(nlcdClass), nrows = 3)
print(facets.plot)
```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: 