---
title: "Assignment 6: GLMs week 1 (t-test and ANOVA)"
author: "Masha Edmondson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on t-tests and ANOVAs. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 18 at 1:00 pm.

## Set up your session

1. Check your working directory, load the `tidyverse`, `cowplot`, and `agricolae` packages, and import the NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv dataset. 

2. Change the date column to a date format. Call up `head` of this column to verify.


```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(agricolae)
library(cowplot)

NTL_LTER.Nutrients <- read_csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

#2
NTL_LTER.Nutrients$sampledate <- as.Date(NTL_LTER.Nutrients$sampledate , format = "%Y-%m-%d")
head(NTL_LTER.Nutrients)
NTL_LTER.Nutrients$year4 <- as.numeric(NTL_LTER.Nutrients$year4)

```


## Wrangle your data

3. Wrangle your dataset so that it contains only surface depths and only the years 1993-1996, inclusive. Set month as a factor.
```{r}
Depths.Totals <- NTL_LTER.Nutrients %>%
 filter(year4 %in% c("1993", "1994", "1995", "1996"), depth_id == "1")

NTL_LTER.Nutrients$month <- as.factor(NTL_LTER.Nutrients$month)
Depths.Totals$month<- as.factor(Depths.Totals$month)
class(Depths.Totals$month)
```

## Analysis

Peter Lake was manipulated with additions of nitrogen and phosphorus over the years 1993-1996 in an effort to assess the impacts of eutrophication in lakes. You are tasked with finding out if nutrients are significantly higher in Peter Lake than Paul Lake, and if these potential differences in nutrients vary seasonally (use month as a factor to represent seasonality). Run two separate tests for TN and TP.

4. Which application of the GLM will you use (t-test, one-way ANOVA, two-way ANOVA with main effects, or two-way ANOVA with interaction effects)? Justify your choice.

> Answer: We would use the ANOVA GLM application to test for nitrogen and phosphorus levels between the two lakes over a three year time period. We would pick two-way anova with interaction because we have two varibales to consider: TN = Lake + month + error and TP= lake + month + error. We are are using ANOVA with interaction because we want to compare the difference in nutrient compositions between the two lakes over seasons.Though the lakes might be similar one is being dumped with nutrients, so the interaction might be able to determine that. We would not use T-test because it could only tell us one predictable variable. We would not use the one-way ANOVA because we are testing for more than one variable and we would not use the two-way ANOVA with main effects because we need to compare the interactions between the lakes.

5. Run your test for TN. Include examination of groupings and consider interaction effects, if relevant.

6. Run your test for TP. Include examination of groupings and consider interaction effects, if relevant.

```{r}
#5
# Format as aov
TN.anova.2way <- aov(data = Depths.Totals, tn_ug ~ month * lakename) 
summary(TN.anova.2way) 
# This summary tells us that the only significant difference is between Peter and Paul lake. Therefore, there is no need to proceede with more tests. Peter lake has the higher concentration of 384 versus Paul lake with only 300.

#6
# Format as aov
TP.anova.2way <- aov(data = Depths.Totals, tp_ug ~ month * lakename) 
summary(TP.anova.2way) #sig difference between lake name and month and we want to inperpret interaction 

# Format as lm
TP.anova.2way2 <- lm(data = Depths.Totals, tp_ug ~ month * lakename)
summary(TP.anova.2way2)

TukeyHSD(TP.anova.2way)

TP_interaction.lake.month <-with(Depths.Totals, interaction(month, lakename))
TP.interaction.anova <- aov(data = Depths.Totals, tp_ug ~ TP_interaction.lake.month)

TP_interaction_groups <-HSD.test(TP.interaction.anova, "TP_interaction.lake.month", group = TRUE)
TP_interaction_groups
```

7. Create two plots, with TN (plot 1) or TP (plot 2) as the response variable and month and lake as the predictor variables. Hint: you may use some of the code you used for your visualization assignment. Assign groupings with letters, as determined from your tests. Adjust your axes, aesthetics, and color palettes in accordance with best data visualization practices.

8. Combine your plots with cowplot, with a common legend at the top and the two graphs stacked vertically. Your x axes should be formatted with the same breaks, such that you can remove the title and text of the top legend and retain just the bottom legend. 

```{r} 
settheme <- theme_classic(base_size = 14) + 
  theme(axis.text = element_text(color = "black"),  
        legend.position = "top") 

theme_set(settheme)

#7 - change the color so it is not the default and set a theme for the graph
TN.anova.plot <- ggplot(Depths.Totals, aes(y = tn_ug, x = month, fill= lakename)) +
  geom_boxplot()+
  scale_fill_viridis_d(begin = 0.4, end = 1.0)+
  labs(x = "Month", y = expression(paste("Total Nitrogen (", mu, "g/L)"))) +
  ylim(0, 2100)
print(TN.anova.plot)

TP.anova.plot <- ggplot(Depths.Totals, aes(y = tp_ug, x = month, fill = lakename)) +
  geom_boxplot()+
  scale_fill_viridis_d(begin = 0.4, end = 1.0)+
  labs(x = "Month", y = expression(paste("Total Phosphorus (", mu, "g/L)")))+
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5, 
               label = c("cd", "bcd", "d", "ab", "d", "a", "d", "abc", "cd", "abcd"), 
               show.legend= FALSE,
               position = position_dodge2(0.6))+
  ylim(0, 70)
print(TP.anova.plot)

#8
plot_grid(TN.anova.plot + theme(axis.text.x = element_blank(),
                                axis.ticks.x= element_blank(),
                                axis.title.x = element_blank()),
          TP.anova.plot + theme(legend.position = "none"),
                                 align = "v", nrow = 2, ncol=1)


```




